COVID19 - Forecast analysis

### 2020-03-13 draft_analysis_Italy.Rmd
Add home page, navigation and theme

---
title: "COVID19 - Forecast analysis"
author: "PG"
date: "3/11/2020"
output: github_document
---

@@

The present analysis used the dataset on COVID19 updated in https://github.com/pcm-dpc/COVID-19.
We used a SEIR model to estimate the prediction of this epidemic in Italy.
We estimated the R0 parameter by means of a simple linear regression ad in https://kingaa.github.io/clim-dis/parest/parest.html

```{r  plot data, echo=TRUE}
plot(dat_csv$data,dat_csv$totale_attualmente_positivi,ylab="Total Covid cases",xlab="Date")
mean(diff(log(dat_csv$totale_attualmente_positivi)))
```
The grow is exponential (rate of increase of about 25\%). 
We estimate the R0 parameter by means of a linear model. R0 indicates how contagious an infectious disease is. It's also referred to as the reproduction number.

@@

where $`Y_t`$ is the cumulative number of infected at the time t, while b is beta, the slope of the regression line.  
We calculate the R0 values based on a different number of days before the last. 
The R0 shows a decreasing trend, but it is stable in the last 12 days.

The slope b indicate the rate of exponetial increase, used for the R0 calculation.   
R code:  
fit1 <- lm(log(totale_attualmente_positivi)~t,data=dat_csv)

```{r data}
head(dat_csv)
fit1 <- lm(log(totale_attualmente_positivi)~t,data=dat_csv)
summary(fit1)
```
The prediction is  good ($R^2$ near 1).
However the tendency is for a reduction of the slope.

```{r model plot, echo=TRUE}
plot(dat_csv$t,log(dat_csv$totale_attualmente_positivi),ylab="log cases",xlab="time")
abline(coef(summary(fit1))[,1])
```
The slope coefficient estimated in the linear regression model can be used to estimate R0.

@@

The incubation period for the coronavirus is in mean 5.1 days with a range from 2-14 days.
Please see https://www.worldometers.info/coronavirus/coronavirus-incubation-period/.
In the calculation we considered an incubation period of 14 days for two reasons: 1) the majority of cases is asymptomatic, contagiousness is greater than 5, maybe 14. A minority (who made the swab) will have a duration of about 5 days between the start of contagiousness and swab; 2) 14 days is the worst scenario to consider.
```{r r0 plot, echo=TRUE}
slope <-coef(summary(fit1))[2,1]; slope
slope.se <- coef(summary(fit1))[2,2]; slope.se
### R0 estimates and 95% IC 
R_0=slope*14+1;R_0
(slope+c(-1,1)*1.96*slope.se)*14+1
```
We want to make a short term forecast (14 days) with 3 scenario:  
-Scenario 1: 10 exposed people for each COVID-19 case and beta the same (no restrictions made or even no effects)
-Scenario 2: 10 exposed people for each COVID-19 case and beta reduced of 50% (-50% exposed people)
-Scenario 3: 5 exposed people for each COVID-19 case and beta reduced of 50%  (-50% both exposed people and COVID19 contagious power)

We fix a series of initial parameters:
-I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0  
- N: Italian population   
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (half of flu epidemic)  
- mu0: the overall mortality rate

```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
#beta 
beta0<-R_0/(14)
# italian poulation
N=60480000
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics)
```{r  scenario plot, echo=TRUE}
library(EpiDynamics)

@@

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$results$I[-1]*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$results$I[-1]*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$results$I[-1]*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```
The 3 scenarios show how measures of restriction can help to reduce the number of infected.  
At the end of the 2 weeks (`r date[length(date)]`) the number of infected is  (`r (seir2$results$I*N)[length(seir2$results$I)]`).
In the next plot the cumulative number of infected.
```{r  cumulative plot, echo=TRUE}

plot(date,c(dat_csv$totale_casi,(seir1$results$I[-1]+seir1$results$R[-1])*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$results$I[-1]+seir2$results$R[-1])*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$results$I[-1]+seir3$results$R[-1])*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
PlotMods(seir2)
```

>>>>

---
title: "COVID19 - Forecast analysis"
author: "PG"
date: "3/13/2020"
output: github_document
---

@@

The present analysis used the dataset on COVID19 updated in https://github.com/pcm-dpc/COVID-19.
We used a SEIR model to predict the trend of the actual COVID19 epidemic in Italy.
We estimated the R0 parameter by means of a linear regression model as reported in https://kingaa.github.io/clim-dis/parest/parest.html

```{r  plot data, echo=TRUE}
plot(dat_csv$data,dat_csv$totale_attualmente_positivi,ylab="Total Covid cases",xlab="Date")
mean(diff(log(dat_csv$totale_attualmente_positivi)))
```
The plot shows an exponential gros. We estimate the R0 parameter by means of a linear model. R0 indicates how contagious an infectious disease is. It's also referred to as "the reproduction number" of COVID19.

@@

where $`Y_t`$ is the cumulative number of infected at the time t, while b is beta, the slope of the regression line.  
We calculate several R0 values, each one based on a different number of days before the last day. 
The R0 shows a decreasing trend in the last period. The slope b indicates the rate of exponetial increase.  
However we considere the value of the last 10 days, because is more stable.
R code:  
```{r data}
head(dat_csv)
fit1 <- lm(log(totale_attualmente_positivi)~t,data=dat_csv[(days-10):days,])
summary(fit1)
```

The slope coefficient estimated in the linear regression model can be used to estimate R0.

@@

The incubation period for the coronavirus is in mean 5.1 days with a range from 2-14 days.
Please see https://www.worldometers.info/coronavirus/coronavirus-incubation-period/.
In the calculation we considered an incubation period of 14 days for two reasons: 1) the majority of cases is asymptomatic, contagiousness is greater than 5, maybe 14. A minority (who made the swab) will have a duration of about 5 days between the start of contagiousness and swab; 2) 14 days is the worst scenario to consider in this phase.
```{r r0 plot, echo=TRUE}
slope <-coef(summary(fit1))[2,1]; slope
slope.se <- coef(summary(fit1))[2,2]; slope.se
### R0 estimates and 95% IC 
R_0=slope*14+1;R_0
(slope+c(-1,1)*1.96*slope.se)*14+1
```
We want to make a short term forecast (14 days) with 3 scenario:  
-Scenario 1: 10 exposed people for each COVID-19 case and beta the same (no restrictions made or even no effects)
-Scenario 2: 10 exposed people for each COVID-19 case and beta reduced of 50% (-50% exposed people)
-Scenario 3: 5 exposed people for each COVID-19 case and beta reduced of 50%  (-50% both exposed people and COVID19 contagious power)

We fix a series of initial parameters:
- I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0  
- N: Italian population   
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (half of flu epidemic)  
- mu0: the overall mortality rate

```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
#beta 
beta0<-R_0/(14)
# italian poulation
N=60480000
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics) and the function SEIR() to implement a SEIR model:

<img src="http://www.public.asu.edu/~hnesse/classes/seireqn.png"/>

```{r  scenario plot, echo=TRUE}
library(EpiDynamics)

@@

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$results$I[-1]*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$results$I[-1]*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$results$I[-1]*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

The 3 scenarios show different numbers. 
If we consider the second scenario, at the end of the 2 weeks (`r date[length(date)]`) the number of infected is  (`r (seir2$results$I*N)[length(seir2$results$I)]`).

In the next plot the cumulative number of infected.  
At the end of the 2 weeks (`r date[length(date)]`) the total number of COVID19 cases is expected to be  (`r ((seir2$results$I+seir3$results$R)*N)[length(seir2$results$I)]`).

```{r  cumulative plot, echo=TRUE}

plot(date,c(dat_csv$totale_casi,(seir1$results$I[-1]+seir1$results$R[-1])*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$results$I[-1]+seir2$results$R[-1])*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$results$I[-1]+seir3$results$R[-1])*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

We make an effort to estimate the effect of restriction measures took by the Italian Government (11 March 2020). 
After 14 days  (`r date[length(date)]`) we have the current status in the second scenario:  
- I14=(`r (seir2$results$I*N)[length(seir2$results$I)]`);  
- R14=(`r (seir2$results$R*N)[length(seir2$results$R)]`);  
- S14=(`r (seir2$results$S*N)[length(seir2$results$S)]`);  
- E14=(`r (seir2$results$E*N)[length(seir2$results$E)]`);

We reduce R0 to 1.10 and the number of exposed to the number of infected.
We estimate the trend of the number infected, recovered and the total cases for other 14 days forward.

```{r  scenario plot2, echo=TRUE}
I14=seir2$results$I[length(seir2$results$I)]
R14=seir2$results$R[length(seir2$results$I)]
S14=seir2$results$S[length(seir2$results$I)]
E14=seir2$results$E[length(seir2$results$I)]
forecast2<-14
beta14<-1.1/14
parameters <- c(mu = mu0, beta = beta14, sigma = sigma0, gamma = 1/duration)
initials <- c(S = 1-I14*2-R14, E = I14, I = I14, R = R14)
seir2_2 <- SEIR(pars = parameters, init = initials, time = 0:forecast2)

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast2+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir2$results$I[-1]*N,seir2_2$results$I[-1]*N),type="l",ylab="Cases",xlab="time",ylim=c(0,((seir2_2$results$I+seir2_2$results$R[-1])*N)[forecast2]),col=2)
lines(date,c(dat_csv$dimessi_guariti,seir2$results$R[-1]*N,seir2_2$results$R[-1]*N),col=3)
lines(date,c(dat_csv$totale_casi,(seir2$results$I[-1]+seir2$results$R[-1])*N,c(seir2_2$results$I[-1]+seir2_2$results$R[-1])*N),col=1)

legend("topleft",c("All cases","Infected","Recovered"),lty=1,col=1:3)
```

### 2020-03-13 draft_analysis_Veneto.Rmd
Add home page, navigation and theme

---
title: "COVID19 - Forecast analysis"
author: "PG"
date: "3/11/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
rm(list=ls())
###import italian dataset updated 10 March 2020
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv",header=T)
dat_csv<-dat_csv[dat_csv$codice_regione==5,]
dat_csv$t<-1:dim(dat_csv)[1]
plot(dat_csv$data,dat_csv$totale_attualmente_positivi,ylab="Infected",xlab="Date")
```

@@

We fix a series of initial parameters:
-I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0  
- N: Italian population   
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (half of flu epidemic)  
- mu0: the overall mortality rate  
```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
#beta 
beta0<-R_0/(14)
# Veneto population
N=4905000
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics)
```{r first scenario plot, echo=TRUE}
library(EpiDynamics)
forecast<-14
parameters <- c(mu = mu0, beta = beta0, sigma = sigma0, gamma = 1/duration)
f1<-10
initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
seir1 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0, sigma = sigma0, gamma = 1/duration)
f2<-5
initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
seir2 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0*1/2, sigma = sigma0, gamma = 1/duration)
f3<-5
initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
seir3 <- SEIR(pars = parameters, init = initials, time = 0:forecast)

>>>>

---
title: "COVID19 - Forecast analysis"
author: "PG"
date: "3/13/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
rm(list=ls())
###import italian dataset updated 13 March 2020
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv",header=T)
dat_csv<-dat_csv[dat_csv$codice_regione==5,]
dat_csv$t<-1:dim(dat_csv)[1]
plot(dat_csv$data,dat_csv$totale_attualmente_positivi,ylab="Infected",xlab="Date")
```

@@

We fix a series of initial parameters:
-I0: initial number of COVID-19 cases  
- R0: initial number of recovered  
- beta: the quantity connected to R0  
- N: Italian population   
- duration: infection duration of COVID-19   
- sigma0: the coronavirus transmission rate (0.05 is the half of flu epidemic)  
- mu0: the overall mortality rate  
```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
#beta 
beta0<-R_0/(14)
# Veneto population
N=4905854
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics)
```{r first scenario plot, echo=TRUE}
library(EpiDynamics)
forecast<-14
parameters <- c(mu = mu0, beta = beta0, sigma = sigma0, gamma = 1/duration)
f1<-10
initials <- c(S = 0.95, E = (f1*I0/N), I = I0/N, R = R0/N)
seir1 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0, sigma = sigma0, gamma = 1/duration)
f2<-5
initials <- c(S = 0.95, E = (f2*I0/N), I = I0/N, R = R0/N)
seir2 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
parameters <- c(mu = mu0, beta = beta0*1/2, sigma = sigma0, gamma = 1/duration)
f3<-5
initials <- c(S = 0.95, E = (f3*I0/N), I = I0/N, R = R0/N)
seir3 <- SEIR(pars = parameters, init = initials, time = 0:forecast)
